Using an iterated block particle filter via \pkg{spatPomp}

### 2024-10-25 ms.Rnw
ben bolker suggestions incorporated

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
        prompt="R> ",
        continue="+  ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )

@@

We now proceed to carry out a similar analysis for the measles model generated by the \pkg{spatPomp} function \code{he10}.
This is a susceptible-exposed-infected-recovered model for measles transmission, described by \citet{ionides22} and \citet{asfaw23arxiv}.
For this model, exact likelihood evaluation is not available.
However, for a relatively small number of units ($U=4$) the particle filter provide an adequate approximation.
\citet{ionides22} considered fitting this model to data using IBPF, with 20 cities and up to $20\times 13$ parameters.
Here, our task is to focus on a smaller, simulated dataset, estimating fewer parameters in order to assess more clearly whether or not the block approximation is leading to substantial bias.
We choose two large towns (London and Birmingham) and two small towns (Cardiff and Hastings) since we expect that population movement from large towns to small towns is essential to explain disase persistence in small towns.
Large towns can maintain an ongoing epidemic, but, below a critical community size local extinction of the disease is expected during epidemic troughs.

@@

Likelihood evaluation took \Sexpr{myround(as.double(m_lik_eval_time,units="mins"),2)} mins.
Recall that the bias-variance trade-off for likelihood evaluation becomes a tradeoff between two sources of bias for log-likelihood evaluation, due to Jensen's inqeuality.
Table~\ref{tab:m-lik-eval} shows that little likelihood is lost due to the block approximation for small numbers of units.
Indeed, even for a large number of particles, a small block size gives higher log-likelihood estimates.
This is on contrast to the results in Table~\ref{tab:lik-eval} for the spatially correlated random walk example in Section~\ref{sec:bm}.

@@

If IBPF is effectively maximizing the BPF approximation to the likelihood, then situations where BPF has low likelihood evaluation bias may correspond to situations where IBPF has low estimation bias.
A profile likelihood for one paramter is presented to support this, in Figure~\ref{fig:m-prof-R0-plot}.
Here, the evaluation using PF gives a slightly tighter estimate of the profile, but this may be less accurate: PF is a higher variance algorithm, even with $U=4$, and its variance increases as the model becomes increasingly misspecified.
Thus, the profile likelihood estimate may have additional curvature due to increasing variance (and therefore increasing Jensen bias) away from the MLE.

@@

Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
We see that both measures agree on the most problematic observation.
However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution.
CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
Using the notation of \citet{asfaw23arxiv}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
\begin{equation}
w^j_{k,n} = \prod_{u \in {\mathcal B}_k}
            f_{Y_{n,n}|X_{u,n}}
            \big(
              y^*_{u,n}\given X^{P,j}_{u,n} \giventh \theta
            \big).
\end{equation}
The BCLL is
\begin{equation}
\hat \ell_{k,n} = \frac{1}{J}\sum_{j=1}^J w^j_{k,n},
\end{equation}
and the block approximation to the conditional likelihood of $y^*_1,\dots,y^*_{U}$ is
\begin{equation}
\hat\ell_n = \sum_{k=1}^K \hat \ell_{k,n}.
\end{equation}
For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
<<bpfilter-diag-code,eval=F,echo=T,purl=F>>=
bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {
  bpfilter(he10_model,Np=switch(i,10,2000),block_size=1)
}
bpf_ll  <- sapply(bpf_list,function(x)x@block.cond.loglik)
dim(bpf_ll) <- c(dim(bpf_list[[1]]@block.cond.loglik),length(bpf_list))
bcll <- apply(bpf_ll,c(1,2),mean)
bcllv <- apply(bpf_ll,c(1,2),var)
@

@@

Trace plots to check on the convergence of \code{ibpf} are similar to those for other iterated filtering algorithms.
In Figure~\ref{fig:traces}, we show this for the Gaussian correlated random walk example.
In this case, we see replicable convergence among independent searches.

<<traces,fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Trace plots for the correlated Gaussian random walk example.">>=
ibpf_traces <- pomp::melt(lapply(ibpf_mle_searches,pomp:::traces_internal))
ibpf_traces$iteration <- as.numeric(ibpf_traces$iteration)
ggplot(ibpf_traces, aes(x=iteration,y=value,group=.L1,color=factor(.L1))) +
  geom_line() +
  guides(color="none") +
  facet_wrap(~variable,scales="free_y")
@

@@

>>>>

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
        prompt="   ",
        continue="   ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )

@@

We now proceed to carry out a similar analysis for the measles model generated by the \pkg{spatPomp} function \code{he10}.
This is a susceptible-exposed-infected-recovered model for measles transmission, described by \citet{ionides22} and \citet{asfaw23arxiv}.
For this model, exact likelihood evaluation is not available.
However, for a relatively small number of units ($U=4$) the particle filter provides an adequate approximation.
\citet{ionides22} considered fitting this model to data using IBPF, with 20 cities and up to $20\times 13$ parameters.
Here, our task is to focus on a smaller, simulated dataset, estimating fewer parameters in order to assess more clearly whether or not the block approximation is leading to substantial bias.
We choose two large towns (London and Birmingham) and two small towns (Cardiff and Hastings) since we expect that population movement from large towns to small towns is essential to explain disase persistence in small towns.
Large towns can maintain an ongoing epidemic, but below a critical community size local extinction of the disease is expected during epidemic troughs.

@@

Likelihood evaluation took \Sexpr{myround(as.double(m_lik_eval_time,units="mins"),2)} mins.
Recall that the bias-variance trade-off for likelihood evaluation becomes a tradeoff between two sources of bias for log-likelihood evaluation, due to Jensen's inquality.
Table~\ref{tab:m-lik-eval} shows that little likelihood is lost due to the block approximation for small numbers of units.
Indeed, even for a large number of particles, a small block size gives higher log-likelihood estimates.
This is in contrast to the results in Table~\ref{tab:lik-eval} for the spatially correlated random walk example in Section~\ref{sec:bm}.

@@

If IBPF is effectively maximizing the BPF approximation to the likelihood, then situations where BPF has low likelihood evaluation bias may correspond to situations where IBPF has low estimation bias.
A profile likelihood for one parameter is presented to support this, in Figure~\ref{fig:m-prof-R0-plot}.
Here, the evaluation using PF gives a slightly tighter estimate of the profile, but this may be less accurate: PF is a higher variance algorithm, even with $U=4$, and its variance increases as the model becomes increasingly misspecified.
Thus, the profile likelihood estimate may have additional curvature due to increasing variance (and therefore increasing Jensen bias) away from the MLE.

@@

Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
We see that both measures agree on the most problematic observations.
However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution.
CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
Using the notation of \citet{asfaw23arxiv}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
\begin{equation}
w^j_{k,n} = \prod_{u \in {\mathcal B}_k}
            f_{Y_{n,n}|X_{u,n}}
            \big(
              y^*_{u,n}\given X^{P,j}_{u,n} \giventh \theta
            \big).
\end{equation}
The BCLL is
\begin{equation}
\hat \ell_{k,n} = \frac{1}{J}\sum_{j=1}^J w^j_{k,n},
\end{equation}
and the block approximation to the conditional likelihood of $y^*_1,\dots,y^*_{U}$ is
\begin{equation}
\hat\ell_n = \sum_{k=1}^K \hat \ell_{k,n}.
\end{equation}
For our measles spatPomp example, we calculate the BCLL and its variance (BCLLV) as follows:
<<bpfilter-diag-code,eval=F,echo=T,purl=F>>=
bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {
  bpfilter(he10_model,Np=switch(i,10,2000),block_size=1)
}
bpf_ll  <- sapply(bpf_list,function(x)x@block.cond.loglik)
dim(bpf_ll) <- c(dim(bpf_list[[1]]@block.cond.loglik),length(bpf_list))
bcll <- apply(bpf_ll,c(1,2),mean)
bcllv <- apply(bpf_ll,c(1,2),var)
@

@@

Trace plots to check on the convergence of \code{ibpf} are similar to those for other iterated filtering algorithms.
In Figure~\ref{fig:traces}, we show this for the Gaussian correlated random walk example.
In this case, we see replicable convergence among independent searches.
The likelihood rises toward the maximum and then stabilizes in a neighborhood of the maximum while the estimated parameters explore this region.

<<traces,fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Trace plots for the correlated Gaussian random walk example. Constant traces for $\\sigma_1$, $\\tau_1$ and $X_{0,1}$ arise because these parameters are treated as known and not estimated. Note that the values of $\\sigma_1$, $\\tau_1$ and $X_{0,1}$  are shared across units (e.g., there is no $\\sigma_2$) since the \\code{bm2} constructor is called with $\\rho$ as the only unit-specific parameter.">>=
ibpf_traces <- pomp::melt(lapply(ibpf_mle_searches,pomp:::traces_internal))
ibpf_traces$iteration <- as.numeric(ibpf_traces$iteration)
ggplot(ibpf_traces, aes(x=iteration,y=value,group=.L1,color=factor(.L1))) +
  geom_line() +
  guides(color="none") +
  facet_wrap(~variable,scales="free_y")
@

@@

\section*{Acknowledgments}

We are grateful to Ben Bolker for providing useful feedback.
